Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-11-21 ATACCC.tex
editing

\Cref{ATACCC:sec:background} explains how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature.
Next, \cref{ATACCC:sec:hakki} explains \textcite{hakkiOnset}'s application of these methods to the ATACCC data.
However, this analysis did not converge fully.
I modify their observation model in section \cref{ATACCC:sec:observation-modification} to improve the convergence.
This modification is based on an earlier analysis I did with a slightly different model and a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
I convert the model and parameter estimates of the biomarker to a duration of PCR positivity in \cref{ATACCC:sec:duration}.
I will refine this estimate in future chapters before its use in estimating transmission.

@@

The second approach is to approximate the dynamics using general mathematical functions.
Most common is a piecewise model, where the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly~\autocites{clearyUsing,kisslerViral,larremoreTest}; although more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
In these models, parameters are clearly related to features of the biomarker's trajectory, such as the time from first detection to peak viral load, and the viral load at the peak.
In both approaches, random effects are used to represent variation between hosts within the population, capturing the non-independence of longitudinal data from the same host.
This chapter uses the model of \textcite{hakkiOnset}.
This model is approximately piecewise log-linear, however, a slight adjustment is made which means that the model is continuous (see \cref{ATACCC:sec:hakki} for details).

@@

Following much of the literature referenced in this section, I will take a Bayesian approach in this chapter.
In the Bayesian context, random effect models are often referred to as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
$\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu_j$ and $\sigma_j$ being the top level.
More complex models could have more levels, for instance if the individuals are grouped into subpopulations.

@@

The model of \textcite{hakkiOnset} has two components: a viral load model and an observation model.
The viral load model produces a determinstic trajectory for the viral load of an individual, based on the model parameters.
The observation model describes how the PCR test noisly observes this viral load.
It includes the possibility of false positive and negatives, and the limit of detection of the PCR test.
\todo{Make sure limit of detection is introduced previously when I first talk about PCR}
% The limit of detection means that viral loads below a certain threshold are not detected by the PCR test.
% They can be represented as a censoring of the viral load at the limit of detection.

Viral loads exhibit exponential growth followed by exponential decline.
\Textcite{hakkiOnset} use a model which is approximately piecewise linear but continuous.
A continuous viral load trajectory should give better performance when using gradient-based inference methods such as Hamiltonian Monte Carlo (HMC)\todo{cite paper or ref section}.
The model for the viral load at time $t$, $\hat{v}_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
\begin{align}
\hat{v}_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time is continuous and relative to the time that individual $i$ was first tested.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As time moves away from the peak (\ie becomes large in magnitude), the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for a representative viral load trajectory).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.
$\tau_{\max,i}$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
\begin{figure}
  \centering \includegraphics{ATACCC/typical_trajectory}
  \caption[Example viral load model trajectory.]{Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values. The increase in log viral load is approximately linear with a gradient of $a_i$, here 5. The decrease is also linear, with gradient $-b_i$, here -1.4. The peak is at time $\tau_{\max,i}$, here 0, and log viral load $\log v_{\max,i}$, here = 15. The smoothing (lack of linearity) near the peak is visible. \label{ATACCC:fig:viral-load-model}}
\end{figure}

@@

The noisy observations are modelled by assuming each measurement of log viral load has iid Normally-distributed error with mean 0 and standard deviation $\sigma_v$.
The measurements are censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$ where $x_0$ and $\sigma_0^2$ are parameters to be estimated.
These readings are censored is the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with $y_{i,t}=0$ if the test is negative.
Then, its likelihood is:
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} > 0$}\\
    \rho F_N(v_\text{log} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$} \label{ATACCC:eq:hakki-likelihood}
\end{cases}
\end{align}
where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a Normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
The first term on each line is the likelihood contribution from test failures, either a false positive (top line) of false negative (bottom line).
The second term on the first line is the likelihood of the observed log viral load from non-failed test results.
The second term on the bottom line is the likelihood of the negative (censored) tests from non-failed test results.

@@

Biological plausibility requires constraining $\vec{\phi}_i$ to the positive reals.
Each $\phi_i$ is modelled as a log-normal random effects by using the hierarchical prior $\phi_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
$\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the log-parameters.
As the random effects have a log-normal, rather than normal, distribution $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
The covariance matrix $\matr{\Sigma}_\phi$ controls the variability of the population and allows for the parameters to be correlated.
For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
To improve sampling efficiency, the multivariate version of the non-centred parameterisation~\autocites{papaspiliopoulosGeneral,stanReparameterization} is used.
This can be expressed as follows.
\begin{align}
  \vec{\eta}_i &\dist \N(0, \matr{C}_\phi) \\
  \log \vec{\phi}_i &= \vec{\mu}_\phi + \vec{\sigma}_\phi \circ \vec{\eta}_i
\end{align}
where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, $\matr{C}_\phi$ is a correlation matrix, and $\circ$ is element-wise multiplication.
Combining $\matr{C}$ and $\vec{\sigma}_\phi$ allows recovering $\matr\Sigma_\phi$ because $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
Correlation matrices have diagonal elements all 1, therefore this reparameterisation has the same number of parameters as the original model.

@@

\section{Modifying the observation model} \label{ATACCC:sec:observation-modification}

@@

Individual 55 has an extremely noisy set of observations.
This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
Because the individual trajectories is not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
\begin{figure}
  \centering \includegraphics{ATACCC/fit_individual_55}
  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain. Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2. \label{ATACCC:fig:individual-55}}
\end{figure}

>>>>

\Cref{ATACCC:sec:background} explains how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature.
Next, \cref{ATACCC:sec:hakki} explains \textcite{hakkiOnset}'s application of these methods to the ATACCC data and in Section \cref{ATACCC:sec:observation-modification} the observation model of Hakki is modified  to improve convergence.
This modification is based on an earlier analysis I did with a slightly different model and a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
I convert the model and parameter estimates of the biomarker to a duration of PCR positivity in \cref{ATACCC:sec:duration}.
I will refine this estimate in future chapters before its use in estimating transmission.

@@

The second approach is to approximate the trajectories from these models using general mathematical functions.
In these models, the function's parameters are clearly related to features of the biomarker's trajectory, such as the time from first detection to peak viral load, and the viral load at the peak.
Most common is a piecewise model, where the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly~\autocites{clearyUsing,kisslerViral,larremoreTest}; although more flexible functions have also been suggested~\autocites{quiltyQuarantine}.

In both approaches, random effects are used to represent variation between hosts within the population, capturing the non-independence of longitudinal data from the same host.
This chapter uses the model of \textcite{hakkiOnset}.
This model is approximately piecewise log-linear, however, a slight adjustment is made which means that the model is continuous (see \cref{ATACCC:sec:hakki} for details).

@@

Following much of the literature referenced in this section, I will take a Bayesian approach in this chapter.
In the Bayesian context, random effect models are often referred to as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
$\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu$ and $\sigma$ being the top level.
More complex models could have more levels, for instance if the individuals are grouped into subpopulations.

@@

The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
The viral load model produces a deterministic trajectory for the viral load of an individual, based on the model parameters.
The observation model then describes how the PCR testing process produces noisy observations of this viral load, including the possibility of false positive and negatives, and the limit of detection of the PCR test.
\todo{Make sure limit of detection is introduced previously when I first talk about PCR}
% The limit of detection means that viral loads below a certain threshold are not detected by the PCR test.
% They can be represented as a censoring of the viral load at the limit of detection.

Viral load exhibits exponential growth followed by exponential decline.
\Textcite{hakkiOnset} use a model which is approximately piecewise linear but continuous.
A continuous viral load trajectory should give better performance when using gradient-based inference methods such as Hamiltonian Monte Carlo (HMC)\todo{cite paper or ref section}.
The model for the viral load at time $t$, $\hat{v}_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
\begin{align}
v_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time is continuous and relative to the time that individual $i$ was first tested.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As $|\tau - tau_\max|$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for a representative viral load trajectory).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.
$\tau_{\max,i}$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
\begin{figure}
  \centering \includegraphics{ATACCC/typical_trajectory}
  \caption[Example viral load model trajectory.]{Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values. The increase in log viral load is approximately linear with a gradient of $a_i$, here 5. The decrease is also linear, with gradient $-b_i$, here -1.4. The peak is at time $\tau_{\max,i}$, here 0, and log viral load $\log v_{\max,i}$, here = 15. The smoothing (lack of linearity) near the peak is visible. \label{ATACCC:fig:viral-load-model}}
\end{figure}

@@

The noisy observations are modelled by assuming each measurement of log viral load is a realisation of a iid Normally-distributed random variable error with mean 0 and standard deviation $\sigma_v$.
The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$ where $x_0$ and $\sigma_0^2$ are parameters to be estimated.
These readings are censored is the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with $y_{i,t}=0$ if the test is negative.
Then, its likelihood is:
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} > 0$}\\
    \rho F_N(v_\text{lod} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$} \label{ATACCC:eq:hakki-likelihood}
\end{cases}
\end{align}
where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a Normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
The first term on each line is the likelihood contribution from test failures, either a false positive (top line) of false negative (bottom line).
The second term on the first line is the likelihood of the observed log viral load from non-failed test results.
The second term on the bottom line is the likelihood of the negative (censored) tests from non-failed test results.

@@

Biological plausibility requires constraining $\vec{\phi}_i$ to positive real values.
Each $\phi_i$ is modelled as a log-normal random effects by using the hierarchical prior $\phi_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
$\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the log-parameters.
As the random effects have a log-normal, rather than normal, distribution $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
The covariance matrix $\matr{\Sigma}_\phi$ controls the variability of the population and allows for the parameters to be correlated.
For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
To improve sampling efficiency, the multivariate version of the non-centred parameterisation~\autocites{papaspiliopoulosGeneral,stanReparameterization} is used.
This can be expressed as follows.
\begin{align}
  \vec{\eta}_i &\dist \N(0, \matr{C}_\phi) \\
  \log \vec{\phi}_i &= \vec{\mu}_\phi + \vec{\sigma}_\phi \circ \vec{\eta}_i
\end{align}
where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, $\matr{C}_\phi$ is a correlation matrix, and $\circ$ is element-wise multiplication.
Combining $\matr{C}$ and $\vec{\sigma}_\phi$ allows recovering $\matr\Sigma_\phi$ because $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
Correlation matrices have diagonal elements all 1, therefore this reparameterisation has the same number of parameters as the original model.

@@

\section{Modifying the Hakki observation model} \label{ATACCC:sec:observation-modification}

@@

Individual 55 has an extremely noisy set of observations.
This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
Because the individual trajectories are not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
\begin{figure}
  \centering \includegraphics{ATACCC/fit_individual_55}
  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain. Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2. \label{ATACCC:fig:individual-55}}
\end{figure}
